Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-07-11 paper-new/concrete-relational-model.tex
some small changes to relational model section

The functors $U$, $\li$, $\times$, and $\arr$ introduced in the previous section
extend to predomains, error domains, and their relations. The functor $U$ simply
takes the underlying predomain relation of an error domain relation. Given a
relation $c_1 : A_1 \rel A_1'$ and $c_2 : A_2 \rel A_2'$, the relation $c_1
\times c_2 : (A_1 \times A_2) \rel (A_1' \times A_2')$ is defined by $((x, y),
(x', y')) \in (c_1 \times c_2)$ if and only if $(x, x') \in c_1$ and $(y, y')
\in c_2$. Given a relation $c : A \rel A'$ and an error domain relation $d : B
\rel B'$, the relation $c \arr d : (A \arr B) \rel (A' \arr B')$ is the error
domain relation defined by $(f, g) \in (c \arr d)$ if and only if for all $x :
A$ and $x' : A'$ with $(x , x') \in c$ we have $(f(x), g(x')) \in Ud$. The fact
that this relation preserves error and respects $\theta$ follows from the
corresponding properties for $d$.

@@

With the notion of syntactic perturbation defined, we can now make formal the
final aspect of the model: the notion of quasi-represntability of relations.
%
\begin{definition}[quasi-left-representable relations]
Let $(A, M_A, i_A)$ and $(A', M_{A'}, i_{A'})$ be value objects, and let $c : A
\rel A'$ be a predomain relation. We say that $c$ is
\emph{quasi-left-representable} by a predomain morphism $e : A \to A'$ if there
are perturbations $\delle_c \in M_A$ and $\delre_c \in M_{A'}$ such that the
following two squares exist: (1) $\upl$: $e \ltsq{c}{\le_{A'}} i_{A'}(\delre_c)$, and
(2) $\upr$: $i_A(\delle_c) \ltsq{\le_{A}}{c} e$.
\end{definition}
%
Observe that this weakens the previous notion of representability, since under
that definition the perturbations were required to be the identity.
%
We make the analogous definition of quasi-left-representability for error domain
relations $d$.

Likewise, we define quasi-right-representability as follows:
%
\begin{definition}
Let $(B, M_B, i_B)$ and $(B', M_{B'}, i_{B'})$ be computation objects, and let
$d : B \rel B'$ be an error domain relation. We say that $d$ is
\emph{quasi-right-representable} by an error domain morphism $p : B' \multimap B$ if
there are perturbations $\dellp_d \in M_B$ and $\delrp_d \in M_{B'}$ such that
the following two squares exist: 
(1) $\dnl$: $p \ltsq{\le_{B'}}{d} i_{B'}(\delrp_d)$ and
(2) $\dnr$: $i_B(\dellp_d) \ltsq{d}{\le_{B}} p$.
\end{definition}
%
We make the analogous definition of quasi-right-representability for predomain
relations $c$.

@@

The action of $\times$ on objects is as follows. Given $(A_1, M_{A_1}, i_{A_1})$
and $(A_2, M_{A_2}, i_{A_2})$, we define the monoid $M_{A_1 \times A_2} =
M_{A_1} \oplus M_{A_2}$. To construct the homomorphism $i_{A_1 \times A_2}$, we
appeal to the universal property and give homomorphisms $M_{A_1} \to
\morbisimid{A_1 \times A_2}$ and $M_{A_2} \to \morbisimid{A_1 \times A_2}$. The
former applies $i_{A_1}$ in the first component and the identity in the second
component, while the latter does the reverse with $i_{A_2}$.

The action of $\arr$ on objects is as follows. Given $(A, M_A, iA)$ and $(B,
M_B, i_B)$ we define the monoid $M_{A \arr B} = M_A^{op} \oplus M_B$. Then to
construct the interpretation as endomorphisms it suffices to define a
homomorphism $M_A^{op} \to \morbisimid{A \arr B}$ and $M_B \to \morbisimid{A
\arr B}$. The former takes an element $m_A$ and a predomain morphism $f : A \to
UB$ and returns the composition $f \circ i_A(m_A)$, while the latter is defined
similarly by post-composition with $i_B$.

@@

Recall the equivalence rules for type precision given in Section \ref{sec:GTLC},
for example the rule that $(c_i \ra c_o)(c_i' \ra c_o')\equiv (c_ic_i' \ra
c_oc_o')$ To validate this rule in the model, we construct two squares $\id
\ltsqbisim{(c_i \ra c_o)(c_i' \ra c_o')}{(c_ic_i' \ra c_oc_o')} \id$ and $\id
\ltsqbisim{(c_ic_i' \ra c_oc_o')}{(c_i \ra c_o)(c_i' \ra c_o')} \id$. Since the
definition of extensional square allows us to specify intermediate morphisms
that are synchronized in the lock-step error ordering, it will be sufficient to
construct two intensional squares involving perturbations on both sides. This
motivates the definition of \emph{quasi-order-equivalence}.

\begin{definition}[quasi-order-equivalence]\label{def:quasi-order-equivalent}
  Let $c, c' : A \rel A'$. We say that $c$ and $c'$ are
  \textbf{quasi-order-equivalent}, written $c \qordeq c'$, if there exist
  perturbations $\delta^l_1, \delta^l_2 \in \pv_A$ and $\delta^r_1, \delta^r_2
  \in \pv_{A'}$ such that there is a square $\delta^l_1 \ltsq{c}{c'} \delta^r_1$
  and a square $\delta^l_2 \ltsq{c'}{c} \delta^r_2$.
\end{definition}
We make the analogous definition for error domain relations $d, d' : B \rel B'$.

@@

\begin{lemma}
\begin{itemize}
  \item $U(d \comp d') \qordeq U(d) \comp U(d')$
  \item $F(c \comp c') \qordeq F(c) \comp F(c')$
  \item $(c \comp c') \to (d \comp d') \qordeq (c \to d) \comp (c' \to d')$
  \item $(c_1 \comp c_1') \times (c_2 \comp c_2') \qordeq (c_1 \times c_2) \comp (c_1'\times c_2')$
\end{itemize}
\end{lemma}
\begin{proof} See appendix \eric{TODO: make sure this is in the appendix}
\end{proof}

>>>>

The functors $U$, $\li$, $\times$, and $\arr$ introduced in the previous section
act on predomains and error domains and also have actions on their morphisms,
relations, and squares. The functor $U$ simply takes the underlying predomain
relation of an error domain relation. Given a relation $c_1 : A_1 \rel A_1'$ and
$c_2 : A_2 \rel A_2'$, the relation $c_1 \times c_2 : (A_1 \times A_2) \rel
(A_1' \times A_2')$ is defined by $((x, y), (x', y')) \in (c_1 \times c_2)$ if
and only if $(x, x') \in c_1$ and $(y, y') \in c_2$. Given a relation $c : A
\rel A'$ and an error domain relation $d : B \rel B'$, the relation $c \arr d :
(A \arr B) \rel (A' \arr B')$ is the error domain relation defined by $(f, g)
\in (c \arr d)$ if and only if for all $x : A$ and $x' : A'$ with $(x , x') \in
c$ we have $(f(x), g(x')) \in Ud$. The fact that this relation preserves error
and respects $\theta$ follows from the corresponding properties for $d$.

@@

With the notion of syntactic perturbation defined, we can now make formal the
final aspect of the model: the notion of quasi-represntability of relations.
%
\begin{definition}[quasi-left-representable relations]\label{def:quasi-left-representable}
Let $(A, M_A, i_A)$ and $(A', M_{A'}, i_{A'})$ be value objects, and let $c : A
\rel A'$ be a predomain relation. We say that $c$ is
\emph{quasi-left-representable} by a predomain morphism $e : A \to A'$ if there
are perturbations $\delle_c \in M_A$ and $\delre_c \in M_{A'}$ such that the
following two squares exist: (1) $\upl$: $e \ltsq{c}{\le_{A'}} i_{A'}(\delre_c)$, and
(2) $\upr$: $i_A(\delle_c) \ltsq{\le_{A}}{c} e$.
\end{definition}
%
Observe that this weakens the previous notion of representability, since under
that definition the perturbations were required to be the identity.
%
We make the analogous definition of quasi-left-representability for error domain
relations $d$.

Likewise, we define quasi-right-representability as follows:
%
\begin{definition}[quasi-right-represntable relations]\label{def:quasi-right-representable}
Let $(B, M_B, i_B)$ and $(B', M_{B'}, i_{B'})$ be computation objects, and let
$d : B \rel B'$ be an error domain relation. We say that $d$ is
\emph{quasi-right-representable} by an error domain morphism $p : B' \multimap B$ if
there are perturbations $\dellp_d \in M_B$ and $\delrp_d \in M_{B'}$ such that
the following two squares exist: 
(1) $\dnl$: $p \ltsq{\le_{B'}}{d} i_{B'}(\delrp_d)$ and
(2) $\dnr$: $i_B(\dellp_d) \ltsq{d}{\le_{B}} p$.
\end{definition}
%
We make the analogous definition of quasi-right-representability for predomain
relations $c$.

@@

The action of $\times$ on objects is as follows. Given $(A_1, M_{A_1}, i_{A_1})$
and $(A_2, M_{A_2}, i_{A_2})$, we define the monoid $M_{A_1 \times A_2} =
M_{A_1} \oplus M_{A_2}$. To construct the homomorphism $i_{A_1 \times A_2}$, we
appeal to the universal property and give homomorphisms $M_{A_1} \to
\morbisimid{A_1 \times A_2}$ and $M_{A_2} \to \morbisimid{A_1 \times A_2}$. The
former applies $i_{A_1}$ in the first component and the identity in the second
component, while the latter does the reverse with $i_{A_2}$.
% Could state this using the functorial action of \times

The action of $\arr$ on objects is as follows. Given $(A, M_A, iA)$ and $(B,
M_B, i_B)$ we define the monoid $M_{A \arr B} = M_A^{op} \oplus M_B$. Then to
construct the interpretation as endomorphisms it suffices to define a
homomorphism $M_A^{op} \to \morbisimid{A \arr B}$ and $M_B \to \morbisimid{A
\arr B}$. The former takes an element $m_A$ and a predomain morphism $f : A \to
UB$ and returns the composition $f \circ i_A(m_A)$, while the latter is defined
similarly by post-composition with $i_B$.
% Could state this using the functorial action of \arr

@@

Recall the equivalence rules for type precision given in Section \ref{sec:GTLC},
for example the rule that $(c_i \ra c_o)(c_i' \ra c_o')\equiv (c_ic_i' \ra
c_oc_o')$ To validate this rule in the model, we construct two squares $\id
\ltsqbisim{(c_i \ra c_o)(c_i' \ra c_o')}{(c_ic_i' \ra c_oc_o')} \id$ and $\id
\ltsqbisim{(c_ic_i' \ra c_oc_o')}{(c_i \ra c_o)(c_i' \ra c_o')} \id$. Since the
definition of extensional square allows us to specify intermediate morphisms
that are synchronized in the intensional ordering, it will be sufficient to
construct two intensional squares between the relations, with perturbations on
both sides. This motivates the definition of \emph{quasi-order-equivalence}.

\begin{definition}[quasi-order-equivalence]\label{def:quasi-order-equivalent}
  Let $A$ and $A'$ be value objects, and let $c, c' : A \rel A'$ be relations
  between the underlying predomains. We say that $c$ and $c'$ are
  \textbf{quasi-order-equivalent}, written $c \qordeq c'$, if there exist
  syntactic perturbations $\delta^l_1, \delta^l_2 \in M_A$ and $\delta^r_1,
  \delta^r_2 \in M_{A'}$ such that there is a square $\delta^l_1 \ltsq{c}{c'}
  \delta^r_1$ and a square $\delta^l_2 \ltsq{c'}{c} \delta^r_2$.
\end{definition}
Given computation objects $B$ and $B$, we make the analogous definition for
relations $d, d' : B \rel B'$ between the underlying error domains.

@@

\begin{lemma}
The following hold:
\begin{itemize}
  \item $U(d \comp d') \qordeq U(d) \comp U(d')$
  \item $\li(c \comp c') \qordeq \li (c) \comp \li (c')$
  \item $(c \comp c') \to (d \comp d') \qordeq (c \to d) \comp (c' \to d')$
  \item $(c_1 \comp c_1') \times (c_2 \comp c_2') \qordeq (c_1 \times c_2) \comp (c_1'\times c_2')$
\end{itemize}

\end{lemma}
\begin{proof} See appendix \eric{TODO: make sure this is in the appendix}
\end{proof}
